# Impactos de datos sin sentido sobre los analisis de dinamica poblacional

Por: RLT

El valor de los estudios cientificos es que los datos representa una visi칩n suficiente cerca a la realidad. El objetivo es tener una apreciaci칩n de los parametros m치s importante para definir un cierto patr칩n o interacciones. Por concequencia la base de todo estudios is based on the fundamewntal units of data collection.  If the data do not represent the reality of the study then the interperettion of the analysis is likely to be erroneous. Fancy mathematics and analyses will NOT result in better or more realistic interpretation if the data is erroneous.  

In this section we will consider different aspects of population projection matrix analises and different aspects of the data collection or analysis which are problematic. For each of the problematic situation I will show an example and how it impacts the results and analyses.  These "Impacts" are not meant to be inclusive of all possible results of non-sensical data, but an example of not considering these issues and how it may warp the intepretations.  Thus a warning to all population biologist to be weary of these issues.  

***

## Small sample size or rare events

Without a doubt the main topic (although exclusive) for the use of PPM has been to evaluate population projection of rare or endangered species (ref). Consequently, sample size of the study or some of the life stages/ages are reduced. These reduced sample size cause parameter estimates and often non-sensical results.  

***

### No Mortality/Perfect survivorship

Consider a species were data is collected and the transition matrix (Sp1matU) is as follows and fertility matriz is (Sp1matF). Note that none of the individuals died in the adult stage (perfect survivorship).  Consiquently, the population size never decreases or increases after a time period.   Under ALL biologically realistic modelling we would expect populaiton size to sdtart decreasing if a population has no receuitment.   

```{r}
library(DiagrammeR)

Sp1matU <- rbind(
  c(0.0, 0.0, 0.0),
  c(0.5, 0.3, 0.0),
  c(0.0, 0.4, 1.0)
) # transition matrix

Sp1matF <- rbind(
  c(0.0, 0.0, 1.0),
  c(0.0, 0.0, 0.0),
  c(0.0, 0.0, 0.0)
) # fertility matrix

Sp1matA = Sp1matU+Sp1matF # TF matrix
library(Rage)
stages <- c("plantula", "juvenil", "adulto")

plot_life_cycle(Sp1matU, stages=stages, fontsize = 0)
```


Thus if we analize this matrix we assume that none of the individuals in the adult stage die, at any time period!!!. Clearly this is not realistic.  For any species the likelyhood of death at any stage is never zero (although is could very small) and thus our matrix is non-sensicle. Consider a tree species, such a *Sequoia*, it is the likely for large trees of this species it is likely to that survivorship is very high, however never 100%.   

In the following script we show that population changes after 5 -6 periods reminds close to one and does not change eventhough we have not recruits (we use the transition matrix without fertility).  If there are no recruits (fertility matrix) population size should reduce with time.  

```{r}
n=c(5,5,5)

library(popdemo)
library(popbio)
truelambda(Sp1matU)
pop.projection(Sp1matU, n=n)$pop.changes
```


#### Assume a small change in mortality

Note that in this new model although surviviroship is very close to zero (0.995), and consequently population decline in every time period (even if only by a small fraction) is possible. 

```{r}
Sp1matU_2 <- rbind(
  c(0.0, 0.0, 0.0),
  c(0.5, 0.3, 0.0),
  c(0.0, 0.4, 0.995)
) # transition matrix

Sp1matF <- rbind(
  c(0.0, 0.0, 1.0),
  c(0.0, 0.0, 0.0),
  c(0.0, 0.0, 0.0)
) # fertility matrix

Sp1matA_2=Sp1matU_2+Sp1matF

pop.projection(Sp1matU_2, n=n)$pop.changes

```

***

### Irreducibility: No Transitions between stages

In the present scenario, juveniles do not grow to become adults, all juveniles either remain juveniles or die. Note the life cycle figure where no arrows connect juveniles to adults.   

```{r}
library(DiagrammeR)

Sp1matU_NT <- rbind(
  c(0.0, 0.0, 0.0),
  c(0.5, 0.7, 0.0),
  c(0.0, 0.0, 0.995)
) # transition matrix

Sp1matF <- rbind(
  c(0.0, 0.0, 1.0),
  c(0.0, 0.0, 0.0),
  c(0.0, 0.0, 0.0)
) # fertility matrix

Sp1matA_NT = Sp1matU_NT+Sp1matF # TF matrix
library(Rage)
stages <- c("plantula", "juvenil", "adulto")

plot_life_cycle(Sp1matU_NT, stages=stages, fontsize = 0)
```


As shown by @caswell2000matrix and more recently explored by @stott2010reducibility matrices need to be *irreducible*.  The concept of *irreducibility* is associated with the life cycle of the species and the matrix includes the transitions from all stages to all other stages.    The above life cycle figure is missing two important components in the life history of the species, no juveniles grow to become adults and none of the adults produce seedlings (seeds that grow to seedlings).  

An easy way of determining if the matrix is irreducible is to read the follwowing script *isIrreducible* in the package (popdemo). Note that is this case the result is *False* for the above matrix. 

```{r}
isIrreducible(Sp1matU_NT) # for the transiiton matrix, missing a transition stage and fertility
isIrreducible(Sp1matA_NT) # for the TF matrix, missing a transition stage

# Now considering Sp1matA_2

isIrreducible(Sp1matA_2) # includes all transitions and fertility
```


***

### No survivorship

Complete mortality of one or more stages. It is often observed that survivorship of smaller individuals or the first stage of the life cycle of a species is highly risky, where the likelihood of survbivorship is very low. For example, most seeds do not survive to germinate. This is likely the norm in orchids where seed production is very high (sometimes millions seeds in one seed capsule, i.e *Catasetum*; ref), but few are shown to germinate (ref). However this is not limited to orchids, in trees the same pattern may be observed, for example in *Nothofagus pumilio* seedling recruitment was less than 1.5% @torres2015seed. NEED EXAMPLE OF NO SEED OR SEEDLING RECRUITMENTS.   

As it is well known that seed germination in orchids is less than straightforward and many variables may influence seed germination @rasmussen2015germination. .......

In the present life cycle graph, none of the seedling survive or growth to the next stage.  Note the first column in the transition matrix all values are zero.  The population may have started with many (even thousands) of seedling, but none grew to become a juvenile or remained as seedlings.  This results in a matrix that is reducible *isIrreducible* = *FALSE* and consequently does not comply with the requirements needed.  

```{r}
library(DiagrammeR)

Sp1matU_NS <- rbind(
  c(0.0, 0.0, 0.0),
  c(0.0, 0.7, 0.0),
  c(0.0, 0.25, 0.995)
) # transition matrix

Sp1matF <- rbind(
  c(0.0, 0.0, 1.0),
  c(0.0, 0.0, 0.0),
  c(0.0, 0.0, 0.0)
) # fertility matrix

Sp1matA_NS = Sp1matU_NS+Sp1matF # TF matrix
library(Rage)
stages <- c("plantula", "juvenil", "adulto")

plot_life_cycle(Sp1matU_NS, stages=stages, fontsize = 0)

isIrreducible(Sp1matU_NS)
```



***

## Fertility estimates and life cycle

The importance of considering how fertility included in the life cycle is extremely important, and  if erroneously considered may result in results and interpretation that are non-sensicle.  We would show a few example of how these can be included in the life cycle and matrix and result in cause problems.


### Incorrect life cycle and fertility stage

Assume you have a species where you model the life cycle as seedling, juvenile and adult.  Fertility is measured as the number of seeds produced by an adult (mean number of seeds per adult).   After collecting the data calculations it is determined that the mean number of seeds per adult is 11000.  If we add this the transition and fertility matrix belowand evaluate the lambda and graph the population growth with the function *project* we have a lambda of 4.93 and after only 5 time periods the number of adults are above 250,000.  Consequently the population growth rate estimeste would be incorrect and misleading. 

The error is that the value in the fertility matrix should correspond to the first stage of the matrix, in this case the seedling stage.  Thus the top right corner of the fertility matrix should not be the mean number of seeds per plant, but the mean number of seedlings produced by a an adult plant.   

```{r}
library(DiagrammeR)

Sp1matU_Fert <- rbind(
  c(0.0, 0.0, 0.0),
  c(0.3, 0.7, 0.0),
  c(0.0, 0.25, 0.995)
) # transition matrix

Sp1matF <- rbind(
  c(0.0, 0.0, 1100.0),
  c(0.0, 0.0, 0.0),
  c(0.0, 0.0, 0.0)
) # fertility matrix

Sp1matA_Fert = Sp1matU_Fert+Sp1matF # TF matrix
library(Rage)
stages <- c("plantula", "juvenil", "adulto")

lambda(Sp1matA_Fert)

# show change in population size
n
pr <- project(Sp1matA_Fert, vector="n", time=5)
plot(pr) # Note that even after just 5 time periods, the number of adults is larger than 250,000 individuals
```

If we wish to add a seed stage than we have to include the stage in our model and matrix with a new stage *the seed stage*.

```{r}
library(DiagrammeR)

Sp1matU_Fert2 <- rbind(
  c(0.0, 0.0, 0.0, 0.0),
  c(0.0001, 0.0, 0.0, 0.0),
  c(0.0, 0.3, 0.7, 0.0),
  c(0.0, 0.0, 0.25, 0.995)
) # transition matrix

Sp1matF2 <- rbind(
  c(0.0, 0.0, 0.0, 1100.0),
  c(0.0, 0.0, 0.0, 0.0),
  c(0.0, 0.0, 0.0, 0.0),
  c(0.0, 0.0, 0.0, 0.0)
) # fertility matrix

Sp1matA_Fert2 = Sp1matU_Fert2+Sp1matF2 # TF matrix
library(Rage)
stages <- c("semillas", "plantula", "juvenil", "adulto")

plot_life_cycle(Sp1matA_Fert2, stages=stages)

lambda(Sp1matA_Fert2)

# show change in population size
pr <- project(Sp1matA_Fert2, vector="n", time=5)
plot(pr) # Note that now the number of adults did not increase to the level of the previous model
```


***


## Data Analysis issues

### Incorrect confidence intervals estimates

  - Confidence intervals of transtion, stasis survivorship based on normal distribution
  
  - Dirichlet approach for estimating confidence intervals
  



